Constraints on Assembly Bias from Galaxy Clustering

### 2016-06-13 ms.tex
Added info on wp measurements, covariances, and likelihoods.

\maketitle
%----------------------------------------------------------------
%%%%%%%%%%%%%%%%%%%%%%%  A B S T R A C T %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We constrain the newly-introduced Decorated HOD model using SDSS DR7
  measurements of projected galaxy clustering, $\wprp,$ and number
  density, $n_{\rm d},$ made from r-band luminosity-threshold
  samples. The Decorated HOD is a model for the galaxy--halo
  connection that augments the traditional Halo Occupation
  Distribution (HOD) by allowing for the possibility of {\em assembly
    bias}: galaxy luminosity may be correlated with dark matter halo
  properties besides virial mass $\mvir$ alone. We demonstrate that it
  is not possible to rule out assembly bias using DR7 measurements of
  $\wprp$ and $n_{\rm d}.$ Moreover, galaxy samples $\magr<-20, -20.5$
  favor strong levels of central galaxy assembly bias:
  high-concentration halos are more likely to host a central galaxy
  relative to low-concentration halos of the same $\mvir.$ We rule out
  zero assembly bias with high significance for these
  samples. Satellite assembly bias only becomes significant for the
  faintest sample we study, $\magr<-19.$ We find no evidence for
  assembly bias in the $\magr<-21$ sample. Because the Decorated HOD
  subsumes the traditional HOD, and additionally because of errors
  made in the original analysis of these samples, our results differ
  from and supersede existing HOD constraints from low-redshift
  clustering measurements.
\end{abstract}

@@

To generate predictions for galaxy clustering, we directly populate dark matter halos with mock galaxies using {\tt Halotools}. 
We explore halo occupation distribution models (HOD) models in this work 
\citep[e.g.][]{seljak00,ma_fry00,scoccimarro01a,berlind02}, though other 
techniques that can be used to interpret such data, such as the conditional luminosity function 
\citep[CLF, e.g.,][]{yang03,vdBosch13}, exist. In this section, we review the ``standard" HOD 
model used in the present work which assumes that there is no galaxy assembly bias. We refer to 
such a model as ``standard" because all HOD analyses of galaxy clustering to date have assumed no 
galaxy assembly bias. In the following section, we describe 
the Decorated HOD model described in \citet{hearin_etal16}. In both cases, we will only 
review the salient features of our methodology briefly; 
interested readers can always refer to {\tt halotools.readthedocs.io} and 
\citet{hearin_etal16} for further details.

@@

All of our analyses are based on the Bolshioi-Planck simulation \citep{riebe_etal11}. 
Bolshoi-Planck was run with cosmological parameters based on \citet{planck13}, $\Omega_{\Lambda} = 0.693, \Omega_{\rm m} = 0.307, \Omega_{\rm b} = 0.048, h = 0.7, {\rm n_s}=0.96$ and $\sigma_8 = 0.82,$ within a cubic box 
$250$ ${\rm Mpc}/h$ on a side, requiring a particle mass of $m_{\rm p} = 1.35\times10^{8}M_{\odot}/h.$ 
Further information about the Bolshoi-Planck simulation is available at \url{https://www.cosmosim.org}.

@@

The five parameters of the standard HOD models are varied in standard analyses are $\log (M_{\rm min})$, 
$\sigma_{\log M}$, $\alpha$, $\log (M_1)$, and $\log (M_0)$, though, as we show below, central galaxies usually 
outnumber satellite galaxies in by a factor of several, so $\log (M_{\rm min})$ and $\sigma_{\log M}$ usually 
vary along a narrow degeneracy that fixes the total galaxy number density to the observed value. There are 
many particular choices that can be made for the functional forms of $\langle N_{\rm cen}\rangle (\mvir)$ and 
$\langle N_{\rm sat}\rangle (\mvir)$. We have made choices that mimic the standard SDSS DR7 analysis of 
\citet{zehavi_etal11}, to expedite comparisons with their results.

@@

After populating a halo catalog with mock galaxies, we calculate the
comoving number density of our mock galaxy sample as $n_{\rm g} =
N_{\rm gal} / L_{\rm box}^3.$ We apply the distant-observer
approximation and use the simulation z-axis as the line-of-sight
direction, and the distance between points in the $xy$-plane to define
the projected distance $\rp.$ We place mock galaxies into
redshift-space by replacing each galaxy's $z$-coordinate with $z
\rightarrow z + V_{\rm z}/H_0.$ We count pairs of points in each of
our $\rp$ bins, rejecting pairs with $z-$distance exceeding $\pi_{\rm
  max}=60$ ${\rm Mpc/h.}$ We employ the \citet{landyszalay93}
estimator to turn pair counts into a $\wprp$ prediction, as described
in detail in {\tt halotools.readthedocs.io}.

@@

In addition to the standard occupation statistics described in the
previous section, in this paper we also use the decorated HOD
formalism to connect galaxies t o dark matter halos in a manner that
has simultaneous dependence on both $\mvir$ and halo concentration.
Briefly, we use Equations \ref{eq:ncen} and \ref{eq:nsat} as our
``baseline" first occupation moments.  At fixed $\mvir,$ halos are
divided into one of two categories, those of high- and
low-concentration, depending on whether the concentration of the halo
places it above or below the rank-order percentile $f_{\rm split},$
which we keep fixed to $f_{\rm split}=0.5$ throughout the paper for
simplicity.  High-concentration halos have a different first
occupation moment relative to low-concentration halos of the same
mass, $\langle\ngal|\mvir, c_{\rm high}\rangle \neq
\langle\ngal|\mvir, c_{\rm low}\rangle.$ The difference between the
first moment of high- and low-concentration halos is modulated by
$\abias,$ the novel parameter of the decorated HOD governing assembly
bias. Values of $\abias=\pm1$ correspond to the maximum strength of
assembly bias allowable by the constraint that the model preserves the
marginalized first moment, $\langle\ngal|\mvir\rangle;$ thus
regardless of the value of $\abias,$ in the decorated HOD the
marginalized first moment of centrals and satellites are {\em
  unchanged from the baseline value defined by Equations \ref{eq:ncen}
  and \ref{eq:nsat}}.  Decorated HOD models all have the same HODs,
when averaged over all halos at fixed mass, as standard HOD
models. The only change in Decorated HOD models is whether or not an
additional property also modulates halo occupation at fixed halo
mass. A value of $\abias = 0$ indicates no galaxy assembly bias
whatsoever. We refer the reader to \citet{hearin_etal16} for further
details about the decorated HOD.

In the present work, we fix our model to the simplest class of galaxy
assembly bias models, though a recipe for generalizing to more
complicated models can be found in \citet{hearin_etal16}. In
particular, we split halos into two populations as specified in the
previous paragraph. We then populate halos with satellite galaxies
specified by an assembly bias parameter $-1 \le A_{\rm sat} \le 1$ and
central galaxies with a distinct assembly bias parameter $-1 \le
A_{\rm cen} \le 1$. These two additional parameters are varied in our
analyses that include assembly bias, making the number of parameters
that vary in these analyses seven. As we show below, these additional
parameters are in many instances poorly constrained by clustering data
of the quality of SDSS DR7 alone, so exploring more complex models of
assembly bias does not yet seem justified.

@@

To infer parameters for the HOD and Decorated HOD models described in the previous subsections, 
we performed a Markov Chain Monte Carlo (MCMC) sampling of the posteriors using the 
affine-invariant ensemble sampler of \citet{goodman_weare10} as implemented in the 
{\tt emcee} software package \citep{foreman-mackey_etal13}. For most cases, we find that 
$\sim 3-10 \times 10^{6}$ samples are necessary in order for our chains to converge.

@@

The most important detail of this analysis is the priors on the parameters. In all analyses 
discussed in this paper, we adopt priors that are uniform distributions over the intervals 
specified in Table~\ref{table:priors}. In the case of the assembly bias parameters 
$A_{\mathrm{cen}}$ and $A_{\mathrm{sat}}$, the priors represent physical 
boundaries. These parameters must satisfy $-1 \le A_{\mathrm{cen,sat}} \le 1$. 
Physical considerations require parameter $\sigma_{\log M} > 0$. All other 
priors have a negligible influence on the posterior aside from $\log M_0$. 
We find that $\log M_0$ is often very poorly constrained by clustering data.

@@

%----------------------------------------------------------------------------------------------------------------------------------
\begin{figure*}
\begin{center}
\includegraphics[width=8.3cm]{Mr19samples.pdf}
\includegraphics[width=8.3cm]{Mr19ABsamples.pdf}
\caption{
{\bf Left:} The $M_r<-19$ threshold sample projected correlation function with diagonal elements of 
covariance (points with errorbars). The grey lines are 25 randomly-selected HOD models that yield 
$\Delta \chi^2 <1$ compared to the best-fitting model. {\bf Right:} Same as the left panel but using a 
fit to a Decorated HOD model that contain parameters to describe the strength of assembly bias.
}
\label{fig:Mr19samples}
\end{center}
\end{figure*}
%---------------------------------------------------------------------------------------------------------------------------------

>>>>

\maketitle
%----------------------------------------------------------------
%%%%%%%%%%%%%%%%%%%%%%%  A B S T R A C T %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We constrain the newly-introduced Decorated HOD model using SDSS DR7
  measurements of projected galaxy clustering, $\wprp,$ and number
  density, $n_{\rm g},$ made from r-band luminosity-threshold
  samples. The Decorated HOD is a model for the galaxy--halo
  connection that augments the traditional Halo Occupation
  Distribution (HOD) by allowing for the possibility of {\em assembly
    bias}: galaxy luminosity may be correlated with dark matter halo
  properties besides virial mass $\mvir$ alone. We demonstrate that it
  is not possible to rule out assembly bias using DR7 measurements of
  $\wprp$ and $n_{\rm g}.$ Moreover, galaxy samples $\magr<-20, -20.5$
  favor strong levels of central galaxy assembly bias:
  high-concentration halos are more likely to host a central galaxy
  relative to low-concentration halos of the same $\mvir.$ We rule out
  zero assembly bias with high significance for these
  samples. Satellite assembly bias only becomes significant for the
  faintest sample we study, $\magr<-19.$ We find no evidence for
  assembly bias in the $\magr<-21$ sample. Because the Decorated HOD
  subsumes the traditional HOD, and additionally because of errors
  made in the original analysis of these samples, our results differ
  from and supersede existing HOD constraints from low-redshift
  clustering measurements.
\end{abstract}

@@

To generate predictions for galaxy clustering, we directly populate dark matter halos with mock galaxies using {\tt Halotools}. 
We explore halo occupation distribution models (HOD) models in this work 
\citep[e.g.][]{seljak00,ma_fry00,scoccimarro01a,berlind02}, though other 
techniques that can be used to interpret such data, such as the conditional luminosity function 
\citep[CLF, e.g.,][]{yang03,vdBosch13}, exist. In this subsection, we review the ``standard" HOD 
model used in the present work, which assumes that there is no galaxy assembly bias. We refer to 
such a model as ``standard" because all HOD analyses of galaxy clustering to date have assumed no 
galaxy assembly bias. In the following subsection, we describe 
the Decorated HOD model described in \citet{hearin_etal16}. In both cases, we will only 
review the salient features of our methodology briefly; 
interested readers can always refer to {\tt halotools.readthedocs.io} and 
\citet{hearin_etal16} for further details.

@@

All of our analyses are based on the Bolshioi-Planck (BolshioP) simulation \citep{riebe_etal11}. 
BolshoiP was run with cosmological parameters based on \citet{planck13}, 
$\Omega_{\Lambda} = 0.693, \Omega_{\rm m} = 1 - \Omega_{\Lambda} = 0.307, \Omega_{\rm b} = 0.048, 
h = 0.7, {\rm n_s}=0.96$ and $\sigma_8 = 0.82,$ within a cubic box 
$250$ ${\rm Mpc}/h$ on a side, requiring a particle mass of $m_{\rm p} = 1.35\times10^{8}M_{\odot}/h.$ 
Further information about the BolshoiP simulation is available at \url{https://www.cosmosim.org}.

@@

The five parameters of HOD models that are varied in standard analyses are $\log (M_{\rm min})$, 
$\sigma_{\log M}$, $\alpha$, $\log (M_1)$, and $\log (M_0)$, though, as we show below, central galaxies usually 
outnumber satellite galaxies by a factor of several, so $\log (M_{\rm min})$ and $\sigma_{\log M}$ usually 
vary along a narrow degeneracy that fixes the total galaxy number density to the observed value. There are 
many particular choices that can be made for the functional forms of $\langle N_{\rm cen}\rangle (\mvir)$ and 
$\langle N_{\rm sat}\rangle (\mvir)$. We have made choices that mimic the standard SDSS DR7 analysis of 
\citet{zehavi_etal11}, to expedite comparisons with their results.

@@

After populating a halo catalog with mock galaxies, we calculate the
comoving number density of our mock galaxy sample as $n_{\rm g} =
N_{\rm gal} / L_{\rm box}^3$, where $N_{\rm gal}$ is the total number of 
galaxies in the mock sample. We apply the distant-observer
approximation and use the simulation z-axis as the line-of-sight
direction, and the distance between points in the $xy$-plane to define
the projected distance $\rp.$ We place mock galaxies into
redshift-space by replacing each galaxy's $z$-coordinate with 
$z_{\rm RS} = z + V_{\rm z}/H_0.$ We count pairs of points in each of
our $\rp$ bins, rejecting pairs with $z_{\rm RS}-$ distance exceeding 
$\pi_{\rm max}=60$ ${\rm Mpc/h}$, which is the same projection 
depth chosen by \citet{zehavi_etal11}. We employ the \citet{landyszalay93}
estimator to turn pair counts into a $\wprp$ prediction, as described
in detail in {\tt halotools.readthedocs.io}.

@@

In addition to the standard occupation statistics described in the
previous section, in this paper we also use the decorated HOD
formalism to connect galaxies to dark matter halos in a manner that
has simultaneous dependence on both $\mvir$ and {\em halo concentration}.
Briefly, we use Equations \ref{eq:ncen} and \ref{eq:nsat} as our
``baseline" first occupation moments.  At fixed $\mvir,$ halos are
divided into one of two categories, those of high- and
low-concentration, depending on whether the concentration of the halo
places it above or below the rank-order percentile $f_{\rm split},$
which we keep fixed to $f_{\rm split}=0.5$ throughout the paper for
simplicity. High-concentration halos have a different first
occupation moment relative to low-concentration halos of the same
mass, $\langle\ngal|\mvir, c_{\rm high}\rangle \neq
\langle\ngal|\mvir, c_{\rm low}\rangle.$ The difference between the
first moment of high- and low-concentration halos is modulated by
$\abias,$ the novel parameter of the decorated HOD governing assembly
bias. Values of $\abias=\pm1$ correspond to the maximum strength of
assembly bias allowable by the constraint that the model preserves the
marginalized first moment, $\langle\ngal|\mvir\rangle;$ thus
regardless of the value of $\abias,$ in the decorated HOD the
marginalized first moment of centrals and satellites are {\em
  unchanged from the baseline value defined by Equations \ref{eq:ncen}
  and \ref{eq:nsat}}. In other words, {\em decorated HOD models all have the same HODs,
when averaged over all halos at fixed mass, as standard HOD
models}. The only change in Decorated HOD models is whether or not an
additional property also modulates halo occupation at fixed halo
mass. A value of $\abias = 0$ indicates no galaxy assembly bias
whatsoever. We refer the reader to \citet{hearin_etal16} for further
details about the decorated HOD.

In the present work, we fix our model to the simplest class of galaxy
assembly bias models, though a recipe for generalizing to more
complicated models can be found in \citet{hearin_etal16}. In
particular, we split halos into two populations as specified in the
previous paragraph. We then populate halos with satellite galaxies
specified by an assembly bias parameter $-1 \le A_{\rm sat} \le 1$ and
central galaxies with a distinct assembly bias parameter 
$-1 \le A_{\rm cen} \le 1$. These two additional parameters are varied in our
analyses that include assembly bias, making the number of parameters
that vary in these analyses seven. As we show below, these additional
parameters are in many instances poorly constrained by clustering data
of the quality of SDSS DR7 alone, so exploring more complex models of
assembly bias does not yet seem justified in such analyses.

@@

We constrain HOD parameters based on SDSS DR7 measurements of the projected 
galaxy two-point functions, $w_{\rm p}(r_{\rm p})$, and galaxy number densities, $n_{\rm g}$, 
for luminosity threshold samples published in \citet{zehavi_etal11}. We use the full covariance 
matrix of the projected correlation function available at 
{\tt astroweb.cwru.edu/izehavi/dr7\_covar/table8}. We assume a likelihood of the 
form $\mathcal{L} \propto e^{-\chi^2/2}$, where
%
\begin{equation}
\label{eq:chisquare}
\chi^2 = \Delta w_{{\rm p},i} \, [C^{-1}]_{ij} \, \Delta w_{{\rm p},j} + \frac{(n_{\rm g}^{\rm mock} - n_{\rm g}^{\rm meas})^2}{\sigma_{\rm n}^2}, 
\end{equation}
%
$\Delta w_{{\rm p},i} = w_{\rm p}^{\rm mock}(r_{{\rm p},i}) - w_{\rm p}^{\rm meas}(r_{{\rm p},i})$ is the difference 
between the projected two-point function predicted by the mock catalog and the measured value in the i$^{\rm th}$ 
separation bin (of 12 bins total), $C^{-1}$ is the inverse of the covariance matrix of the measurements, 
and the last term in Eq.~(\ref{eq:chisquare}) is the contribution from the difference between the predicted and measured 
galaxy number density. The error on the galaxy number density assumes Poisson statistics for both the measured and predicted 
galaxy number densities.

To infer parameters for the HOD and Decorated HOD models described in the previous subsections, 
we performed a Markov Chain Monte Carlo (MCMC) sampling of the posterior distribution using the 
affine-invariant ensemble sampler of \citet{goodman_weare10} as implemented in the 
{\tt emcee} software package \citep{foreman-mackey_etal13}. For most cases, we find that 
$\sim 3-10 \times 10^{6}$ samples are necessary in order for our chains to converge.

@@

The most important detail of this analysis is the priors on the parameters. In all analyses 
discussed in this paper, we adopt priors that are uniform distributions over the intervals 
specified in Table~\ref{table:priors}. In the case of the assembly bias parameters 
$A_{\mathrm{cen}}$ and $A_{\mathrm{sat}}$, the priors represent physical 
boundaries. These parameters must satisfy $-1 \le A_{\mathrm{cen,sat}} \le 1$. 
Physical considerations require parameter $\sigma_{\log M} > 0$. All other 
priors have a negligible influence on the posterior aside from $\log M_0$. 
We find that $\log M_0$ is often very poorly constrained by clustering data and 
priors on $\log M_0$ can have a non-negligible influence on inferred parameters.

@@

%----------------------------------------------------------------------------------------------------------------------------------
\begin{figure*}
\begin{center}
\includegraphics[width=8.3cm]{Mr19samples.pdf}
\includegraphics[width=8.3cm]{Mr19ABsamples.pdf}
\caption{
{\bf Left:} The $M_r<-19$ threshold sample projected correlation function with diagonal elements of 
covariance (points with errorbars). The grey lines are 50 randomly-selected HOD models that yield 
$\Delta \chi^2 <1$ compared to the best-fitting model. {\bf Right:} Same as the left panel but using a 
fit to a Decorated HOD model that contain parameters to describe the strength of assembly bias.
}
\label{fig:Mr19samples}
\end{center}
\end{figure*}
%---------------------------------------------------------------------------------------------------------------------------------
